An Accurate Model for SARS-CoV-2 pooled RT-PCR Test Errors

### 2021-03-30 comment.tex
blocking arrow, remove paragraphs

\author[1,2]{Yair Daon, PhD}
\author[2,3]{Amit Huppert, PhD}
\author[1,2]{Uri Obolski, PhD}

@@

\affil[3]{The Gertner Institute for Epidemiology and Health Policy
  Research, Tel Hashomer, Israel}
%\email{yair.daon@gmail.com}
\date{}

@@

Formally, we consider a pool containing $N$ individuals
$\{1,\dots,N\}$. We denote the true infection states $\theta \in
\{0,1\}^N$, so individual $i$ is infected iff $\theta_i=1$. The RT-PCR
test's sensitivity (true-negative rate) is denoted $\sens$, and the
test's specificity (true-positive rate) is denoted $\spec$. Pooled
test result (data) is denoted $\data$, where $\data=0$ iff the test
returned a negative result.

@@

\subsection*{Refuting the common assumption}\label{subsec:refute}
We refute the common assumption with experimental data summarized in
Table \ref{table}, collected from \cite{Salazar}. There, the authors
investigate Dorfman pooling and, regardless of the pooled test result,
follow up and test each pool member separately. We focus on 128 pools
for which at least one subsequent separate test was positive --- of
which 29 pooled tests were negative and 99 positive. In the data cited
in \cite{Salazar}, of the 29 negative pools, subsequent separate
testing yielded a single positive result in 24. In contrast, of the 99
positive pools, 42 yielded a single positive test upon subsequent
separate testing.

The data in Table \ref{table} allows us to test the following null
hypothesis $H_0$: The probability of a pooled false-negative is equal
for pools with one subsequent positively tested member and pools with
two or more such members. $H_0$ is a direct consequence of the
assumptions, and rejecting $H_0$ implies the common assumption are not
realistic, at least for SARS-CoV-2.

@@

\subsection*{Our model}\label{subsec:ours}
Since the essence of the refuted common assumption is that
amplification of all samples occurs only once, we assume the opposite:
amplification of viral RNA occurs (or fails) for each sample
independently. According to \cite{Simplistic1, Simplistic2, Kim,
  OptimalDorfmanPool}, a false-positive does not
depend on the number of negative samples in a pool. We incorporate
this assumption in our model with a small modification: We assume that
an erroneous amplification can occur in \emph{any} pool. Specifically,
it is possible that correct amplifications fails and an erroneous one
occurs simultaneously. This assumption is somewhat specific for the
current application of screening for SARS-CoV-2 via RT-PCR. For
example, cross-reactivity with other coronaviruses would have violated
this assumption. However, cross-reactivity was ruled out in
\cite{KitComparison}.
% To summarize, we assume that for a single pool, a positive test result
% is generated in one of two paths. Either, SARS-CoV-2 RNA from an
% infected individual's sample is correctly amplified and detected, and
% this can happen for each positive sample in a pool, with probability
% $\Se$. Or, some erroneous amplification occurs (e.g. contaminant viral
% RNA is introduced), an event that occurs at most once per pool, with
% probability $1-\Sp$.
These assumptions lead to our model, which is illustrated in 
Figure~\ref{fig:likelihood} and summarized in
%% A negative pooled test result occurs when all detection paths (both
%% correct and erroneous) fail. The probability of no false-detection
%% accounts for the $\spec$ term. The probability of no correct detection
%% is $(1-\sens)$ per infected individual. The probability that all such
%% paths fail is the product of the above mentioned terms, displayed in
\eqref{eq:likelihood a}.%%  The probability of a positive result,
%% presented in \eqref{eq:likelihood b}, is simply the complement.
%% The probability of a negative pooled test result is presented in
%% \eqref{eq:likelihood a} (along with its complement
%% \eqref{eq:likelihood b}), and explained below.
%\begin{subequations}
\begin{align}%\label{eq:single likelihood}
  %\begin{split}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N
    (1-\sens)^{\theta_i} = \spec(1-\Se)^{\sum_i\theta_i}. \label{eq:likelihood a}
    %
    %
    %
    %% \Pr(\data=1 | \theta) &= 1 - \spec\prod_{i=1}^N
    %% (1-\sens)^{\theta_i} = 1 - \spec(1-\Se)^{\sum\theta_i} . \label{eq:likelihood b}
  %\end{split}
\end{align}
%\end{subequations}
%% Combining \eqref{eq:likelihood a} and \eqref{eq:likelihood b}, and
%% recalling that $\data \in \{0,1\}$ yields:

@@

    \begin{scope}[>={Stealth[black]},
        every node/.style={fill=white},
        every edge/.style={draw=black,very thick}]
      \path [->] (A) edge node {$1-\sens$} (2,4.6);
      %\path [->] (B) edge node {$0$} (G);
      %\path [->] (D) edge node {$0$} (G);
      \path [->] (B) edge node {$1-\sens$} (2.5,4.6);
        \path [->] (F) edge node {$\spec$} (3.7,5); 
    \end{scope}
  \end{tikzpicture}
  \caption{Illustration of our model. A pool contains individuals
    $\{1,2,3\}$ with state $\theta=(1,1,0)$ (i.e., individuals 1
    and 2 are infected). A negative pooled test ($\data=0$) occurs when
    three detection paths fail: A false-negative occurs for individuals
    $1$ and $2$, each with probability $(1-\sens)$.
    Additionally, no false-positive detection of SARS-CoV-2 occurs,
    with probability $\spec$. Individual $3$ is not infected
    and does not contribute to the probability of the pooled test
    result. The pooled test is positive if any path results in a detection.}\label{fig:likelihood}
\end{figure}

@@

According to the common assumption, the scheme false-negative rate is
$1-\Se^2$. A short calculation shows that this approximation implies
the percentage increase in scheme false-negative rate is just
$\err{common} := 100\cdot \Se$.

@@

\section*{Discussion}
%\subsection{Short recap}
Here, we presented a novel probabilistic model for the outcomes of pooled PCR tests, parametrized for SARS-CoV-2. Contrary to previous models, we assume, based on data (see section XX), that multiple infected individuals increase the chance of a positive pool.
A direct consequence of our model is that  false-negative rates depend on infection prevalence. Specifically, low values of infection prevalence increase
$\err{exact}$. These results remain qualitatively similar under varying
parameter values, in the observed ranges
\cite{KitComparison,EstimatingRatesKucrika, EstimatingRatesLourenco,
  InterpretingCOVID19Test} (Figure \ref{fig1}).

Our results lead to a conflict: low infection prevalence leads to high efficiency of Dorfman pooling \cite{DorfmanYuvalDor}, while also increasing false-negative rates.

fact almost disregarded in
\cite{DorfmanYuvalDor}: although (Dorfman) pooling is most efficient
when infection prevalence is low, such circumstances are exactly those leading
to a substantial increase in false-negative rates. Such behavior of
pooling is not implied by the common assumption

@@

One important consequence of the efficiency charts is that
practitioners order RT-PCR reagents, they should expect \emph{lower}
scheme efficiency, and thus order more reagents.

%\subsection{Limitations}
Lacking large amounts of experimental data, it is hard to assess the
validity of our model. The data in \cite{Salazar} does allow us to
reject the common assumption but not much more. Our model is as simple
as possible, given that the standard assumptions are refuted and
dilution effects are not taken into consideration.

@@

Another important factor in pooling is dependence between samples. As mentioned in \cite{DorfmanYuvalDor}, introducing a positive
dependence within a pool decreases the false-positive rate. In the
extreme case, consider a fully connected pool, where one infection
determines the entire pool is infected. In this case, a calculation
analogous to the one conducted above (section??) recovers the initial
false-negative rate $1-\Se$. Interestingly, pooling was also noted to
have increased throughput when infection probabilities are dependent
between the pooled individuals \cite{DorfmanYuvalDor}, providing
another advantage to sampling dependent individuals in pooling
schemes. Hence, deliberately pooling individuals with positive dependence of infection should be further studied and likely encouraged when performing pooling.

To conclude, pooling is an important technique that can improve
testing throughput in a cost-effective manner. Nevertheless, care must
be given to pooling schemes' false-negative rates, especially under low infection prevalence settings.

>>>>

\author[1,2]{Yair Daon}
\author[2,3,*]{Amit Huppert}
\author[1,2,*]{Uri Obolski}

@@

\affil[3]{The Gertner Institute for Epidemiology and Health Policy
  Research, Tel Hashomer, Israel}
\affil[*]{Equal contribution}
%\email{yair.daon@gmail.com}
\date{}

@@

Formally, we consider a pool containing $N$ individuals
$\{1,\dots,N\}$. We denote the true infection state $\theta \in
\{0,1\}^N$, so individual $i$ is infected iff $\theta_i=1$. The RT-PCR
test's sensitivity (true-negative rate) is denoted $\sens$, and the
test's specificity (true-positive rate) is denoted $\spec$. Pooled
test result (data) is denoted $\data \in \{0,1\}$, where $\data=0$ iff
the test returned a negative result.

@@

\subsection*{Refuting the common assumption}\label{subsec:refute}
We refute the common assumption with experimental data collected from
\cite{Salazar}, and summarized in Table \ref{table}. There, the
authors investigate Dorfman pooling and, regardless of the pooled test
result, follow up and test each pool member separately. We focus on
128 pools for which at least one subsequent separate test was positive
--- of which 29 pooled tests were negative and 99 positive. In the
data cited in \cite{Salazar}, of the 29 negative pools, subsequent
separate testing yielded a single positive result in 24. In contrast,
of the 99 positive pools, 42 yielded a single positive test upon
subsequent separate testing.

The data in Table \ref{table} allows us to test the following null
hypothesis $H_0$: The probability of a pooled false-negative is equal
for pools with one subsequent positively tested member and pools with
two or more such members. $H_0$ is a direct consequence of the common
assumption, and rejecting $H_0$ implies the common assumption is not
realistic, at least for SARS-CoV-2.

@@

\subsection*{Our model}\label{subsec:ours}
Since the essence of the refuted common assumption is that
amplification of all samples occurs only once, we assume the opposite:
amplification of viral RNA succeeds or fails for each sample
independently. According to \cite{Simplistic1, Simplistic2, Kim,
  OptimalDorfmanPool}, a false-positive does not
depend on the number of negative samples in a pool. We incorporate
this assumption in our model with a small modification: We assume that
an erroneous amplification can occur in \emph{any} pool. Specifically,
it is possible that correct amplifications fails and an erroneous one
occurs simultaneously. This assumption is somewhat specific for the
current application of screening for SARS-CoV-2 via RT-PCR. For
example, cross-reactivity with other coronaviruses would have violated
this assumption. However, cross-reactivity was ruled out in
\cite{KitComparison}.
% To summarize, we assume that for a single pool, a positive test result
% is generated in one of two paths. Either, SARS-CoV-2 RNA from an
% infected individual's sample is correctly amplified and detected, and
% this can happen for each positive sample in a pool, with probability
% $\Se$. Or, some erroneous amplification occurs (e.g. contaminant viral
% RNA is introduced), an event that occurs at most once per pool, with
% probability $1-\Sp$.
These assumptions lead to our model, which is illustrated in 
Figure~\ref{fig:likelihood} and summarized in
%% A negative pooled test result occurs when all detection paths (both
%% correct and erroneous) fail. The probability of no false-detection
%% accounts for the $\spec$ term. The probability of no correct detection
%% is $(1-\sens)$ per infected individual. The probability that all such
%% paths fail is the product of the above mentioned terms, displayed in
\eqref{eq:likelihood a}.%%  The probability of a positive result,
%% presented in \eqref{eq:likelihood b}, is simply the complement.
%% The probability of a negative pooled test result is presented in
%% \eqref{eq:likelihood a} (along with its complement
%% \eqref{eq:likelihood b}), and explained below.
%\begin{subequations}
\begin{align}%\label{eq:single likelihood}
  %\begin{split}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N
    (1-\sens)^{\theta_i} = \spec(1-\Se)^{\sum_i\theta_i}. \label{eq:likelihood a}
    %
    %
    %
    %% \Pr(\data=1 | \theta) &= 1 - \spec\prod_{i=1}^N
    %% (1-\sens)^{\theta_i} = 1 - \spec(1-\Se)^{\sum\theta_i} . \label{eq:likelihood b}
  %\end{split}
\end{align}
%\end{subequations}
%% Combining \eqref{eq:likelihood a} and \eqref{eq:likelihood b}, and
%% recalling that $\data \in \{0,1\}$ yields:

@@

    \begin{scope}[>={Stealth[black]},
        every node/.style={fill=white},
        every edge/.style={draw=black,very thick}]
      \path [->] (A) edge node {$1-\sens$} (2,4.6);
      %\path [->] (B) edge node {$0$} (G);
      %\path [->] (D) edge node {$0$} (G);
      \path [->] (B) edge node {$1-\sens$} (2.5,4.6);
      \path [->] (F) edge node {$\spec$} (3.7,5);
      \path [-Bar] (D) edge (4.5, 2);
    \end{scope}
  \end{tikzpicture}
  \caption{Illustration of our model. A pool contains individuals
    $\{1,2,3\}$ with state $\theta=(1,1,0)$ (i.e., individuals 1 and 2
    are infected). A negative pooled test ($\data=0$) occurs when
    three detection paths fail: A false-negative occurs for
    individuals $1$ and $2$, each with probability $(1-\sens)$.
    Additionally, no false-positive detection of SARS-CoV-2 occurs,
    with probability $\spec$. Individual $3$ is not infected and does
    not contribute to the probability of the pooled test result. The
    pooled test is positive ($\data=1$) if any path results in a
    detection.}\label{fig:likelihood}
\end{figure}

@@

According to the common assumption, the scheme false-negative rate is
$1-\Se^2$. A short calculation shows that this approximation implies
the percentage increase in scheme false-negative rate is $\err{common}
:= 100\cdot \Se$.

@@

\section*{Discussion}
%\subsection{Short recap}
Here, we presented a novel probabilistic model for the outcomes of
pooled PCR tests, parametrized for SARS-CoV-2. Contrary to the common
assumption, we assume, based on data (see section XX), that multiple
infected individuals increase the likelihood of a positive pooled test
result. A direct consequence of our model is that false-negative rates
depend on infection prevalence. Specifically, low values of infection
prevalence increase $\err{exact}$. These results remain qualitatively
similar under varying parameter values, in the observed ranges
\cite{KitComparison,EstimatingRatesKucrika, EstimatingRatesLourenco,
  InterpretingCOVID19Test} (Figure \ref{fig1}).

Our results lead to a conflict: low infection prevalence leads to high
efficiency of Dorfman pooling \cite{DorfmanYuvalDor}, while also
increasing false-negative rates. This is a fact almost disregarded in
\cite{DorfmanYuvalDor}: although (Dorfman) pooling is most efficient
when infection prevalence is low, such circumstances are exactly those
leading to a substantial increase in false-negative rates. Such
behavior of pooling is not implied by the common assumption.

@@

%% One important consequence of the efficiency charts is that
%% practitioners order RT-PCR reagents, they should expect \emph{lower}
%% scheme efficiency, and thus order more reagents.

%\subsection{Limitations}
Lacking large amounts of experimental data, it is hard to assess the
validity of our model. The data in \cite{Salazar} does allow us to
reject the common assumption but not much more. Our model assumes as
little as as possible, given that the standard assumptions are refuted
and dilution effects are not taken into consideration.

@@

%% Another important factor in pooling is dependence between samples. As
%% mentioned in \cite{DorfmanYuvalDor}, introducing a positive dependence
%% within a pool decreases the false-positive rate. In the extreme case,
%% consider a fully connected pool, where one infection determines the
%% entire pool is infected. In this case, a calculation analogous to the
%% one conducted above (section??) recovers the initial false-negative
%% rate $1-\Se$. Interestingly, pooling was also noted to have increased
%% throughput when infection probabilities are dependent between the
%% pooled individuals \cite{DorfmanYuvalDor}, providing another advantage
%% to sampling dependent individuals in pooling schemes. Hence,
%% deliberately pooling individuals with positive dependence of infection
%% should be further studied and likely encouraged when performing
%% pooling.

%% To conclude, pooling is an important technique that can improve
%% testing throughput in a cost-effective manner. Nevertheless, care must
%% be given to pooling schemes' false-negative rates, especially under
%% low infection prevalence settings.
